Free surface computations in mantle convection models

### 2015-07-31 free-surface-paper.tex
more text

\hl{address pseudo free surface, free surface, sticky air}

More recently, the nature of surface boundary conditions have been shown to be important for controlling the 
dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
participating codes in that benchmark used the `sticky air' approximation for the free surface, where a low viscosity,
low density layer of fluid (termed `air', though its viscosity is much higher) is placed above the free surface, decoupling 
it from the boundary. Typically a free-slip boundary condition is used above the stiky air.

\citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it 
introduces.

A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
with a weak crust is important for producing one-sided subduction zones.

Both true free surface and sticky air calculations have been subject to an instability which has been 
variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization}. 
This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
internal density contrasts), severely limits the maximum stable timestep for free surface computations.
Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
equivalent model with free-slip boundary conditions.

@@

Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
The exponentials may be expanded to find
\begin{equation}
a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
\end{equation}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
\caption{Sensitivity to $\tau^*$. As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces. The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the essentially spectral accuracy of the time marching scheme for that case.}
\label{fig:tau_sensitivity}
\end{figure*}

>>>>

There have been several approaches to treating real free surfaces in geodynamic simulations.
\citet{zhong1996free} introduced a pseudo-free-surface formulation, where the free surface coordinate was
treated as an extra variable that was integrated in time. In this approximation, the free surface 
is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface 
corresponding to the hydrostatic pressure perturbation predicted by the surface topography.

A large number of studies have approximated free surfaces in the interior of the domain by using the 
`sticky air' approximation. In this approximation there is a low-viscosity, low-density layer in the fluid 
(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
Typically a free-slip boundary condition is used above the sticky air.

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
of the domain. In this case, there can be flow in an out of the domain, so the boundary must move in time.
A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells.

Recently, the nature of surface boundary conditions have been shown to be important for controlling the 
dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
participating codes in that benchmark used the `sticky air' approximation for the free surface, where a low viscosity,
low density layer of fluid (termed `air', though its viscosity is much higher) is placed above the free surface, decoupling 
it from the boundary. Typically a free-slip boundary condition is used above the stiky air.

\citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it introduces.
A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
with a weak crust is important for producing one-sided subduction zones.

All of the approaches to free surface simulations have been subject to an instability which has been 
variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
internal density contrasts), severely limits the maximum stable timestep for free surface computations.
Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
equivalent model with free-slip boundary conditions.

@@

Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
The exponentials may be expanded to find
\ifdetail
\begin{equation}
a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
\end{equation}
\fi
\begin{equation}
a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3)
\end{equation}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
\caption{Sensitivity to $\tau^*$. As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces. The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the almost spectral accuracy of the time marching scheme for that case.}
\label{fig:tau_sensitivity}
\end{figure*}
